White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-09-21 papers/wdmerger_I/paper.tex
Point out that we don't do source-term predictor when we don't have good old-time data
Used numbered appendix option
More paper I updates in response to comments

\documentclass[iop]{../emulateapj}

@@

\author{Max P. Katz, Michael Zingale, Alan C. Calder, F. Douglas Swesty, Ann S. Almgren, Weiqun Zhang}
%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
The Type Ia supernova progenitor problem is one of the most perplexing and 
exciting problems in astrophysics, requiring detailed numerical modeling to 
complement the observations of these explosions. One possible progenitor that has 
merited recent theoretical attention is the white dwarf merger scenario.
This can naturally explain many of the observed characteristics of 
Type Ia supernovae, as well as their distribution in time and space.
To date there have been few fully self-consistent simulations 
of merging binary white dwarf systems using mesh-based hydrodynamics, 
relative to other methods. This is the first paper in a series designed to 
describe simulations of these systems using a mesh-based hydrodynamics code with 
adaptive-mesh refinement. In this paper we describe our numerical 
methodology and discuss our implementation in the code \castro. \castro\ 
solves the Euler equations 
and the Poisson equation for self-gravity. Standard techniques for 
coupling gravitation and rotation forces to the hydrodynamics do 
not do an adequate job conserving the total energy of the system, 
but recent advances in the literature allow progress on this 
problem and we discuss our implementation here. We present a
set of test problems demonstrating the extent to which our software can sufficiently
model a system where large amounts of mass are advected on the computational 
domain over very long timescales. Future papers in this series will describe
our treatment of the initial conditions of these systems and will 
examine the early phases of the merger to determine its viability
for triggering a thermonuclear detonation.

@@

Around the same time, however, \cite{guillochon:2010} and
\cite{dan:2011} pointed out that the previously mentioned simulations 
generally shared a significant drawback, which was that their initial conditions
were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006},
and \cite{motl:2007} (the first three-dimensional grid-based
simulations of mass transfer in binary white dwarf systems) pioneered
the study of looking at the long-term dynamical evolution of binary
white dwarf systems after constructing equilibrium initial
conditions. Earlier work placed the stars too close together 
and ignored the effects of tidal forces that change the shape of the 
secondary, leading to the merger
happening artificially too quickly \citep{fryer:2008}. When the initial conditions are
constructed in equilibrium, the system can be stable for tens of
orbital periods, substantially changing the character of the mass
transfer phase. One limitation of this series of studies is
that the authors used a polytropic equation of state and thus could
not consider nuclear reactions. \cite{guillochon:2010} and
\cite{dan:2011} improved on this using a realistic equation of state,
a nuclear reaction network, and a similar approach to the equilibrium
initial conditions, and found substantial agreement with the idea that
mass transfer occurs in a stable manner over tens of orbital
periods. They also found that, assuming the material accreted onto the
surface of the primary was primarily helium, explosive surface
detonations would occur as a result of accretion stream instabilities
during the mass transfer phase prior to the full merger. This could
trigger a double-detonation explosion and thus perhaps a SN Ia.

The latest developments have resulted in some possible areas of convergence.
\cite{pakmor:2012} performed a merger scenario
with a $1.1\ \msolar$ and $0.9\ \msolar$ setup, with better treatment
of the initial conditions, and indeed found that the merger process
happened over more than ten orbits. Nevertheless, they still determined
that a carbon-oxygen detonation would occur, in line with their
earlier results. \cite{moll:2014} and \cite{kashyap:2015} were also 
able to find a detonation in similarly massive systems. Notably,
the detonation occurred self-consistently and did not need to be inserted 
by hand. \cite{dan:2012} and \cite{dan:2014} performed a large sweep 
of the parameter space for merger pairs and
found that pure carbon-oxygen systems would generally not lead to
detonations and violent mergers except for the most massive
systems. They did find that for systems containing helium, many
would detonate and potentially lead to SNe Ia, either through the
aforementioned instabilities in the accretion stream, or during the
contact phase, similar to the violent carbon-oxygen
mergers. \cite{pakmor:2013} added a thin helium shell on their primary
white dwarf, and found that this robustly led to a detonation of the
white dwarf. For now there is preliminary support for the hypothesis
that systems with helium shells (or helium WDs), and very massive carbon-oxygen binaries,
could robustly lead to events resembling SNe Ia.

Given the considerable research into the double degenerate problem 
described above, why is another approach using a different simulation
code warranted? First and foremost, reproducibility of the results
across simulation codes and algorithms is important for gauging
confidence in this result. Most of the existing results that study 
the viability of double degenerate systems as progenitors for
Type Ia supernovae (that is, including a realistic 
equation of state and nuclear reactions) have
used the SPH method. SPH codes have a number of features which do aid
them in the study of these systems, such as excellent conservation of
angular momentum. A drawback relates to the fact that whether a prompt detonation
in a merger happens depends in detail on the nature of the
gas at the interface between the two stars, which is at much lower
density than the rest of the stellar material. The SPH codes use
uniform mass particles, so their effective resolution is
\textit{lowest} at the stellar surface. In contrast, a grid-based code
with adaptive mesh refinement can zoom in on the regions where
hotspots will develop, while also maintaining high enough resolution
in the high-density regions to adequately capture the large-scale mass
transfer dynamics. However, there are also outstanding questions of
convergence in SPH (e.g.\ \citealt{zhu_SPH:2014}) and whether the method
correctly captures fluid instabilities. This is an important question
for white dwarf mergers because of the likely importance small-scale
instabilities will have on the evolution of the low-density gas at the
primary's surface. The pioneering work of \cite{agertz:2007} compared
grid and SPH codes and found some important differences. Most relevant
for this discussion is that the SPH codes could not adequately handle
mixing from the Kelvin-Helmholtz instability in the test they
propose. As pointed out by \cite{price:2008}, this is not a result of
SPH being inherently unable to model this instability, but instead it
is attributed to the fact that the standard SPH evolution equations do
not have a mechanism for capturing discontinuities in internal
energy. \citeauthor{price:2008} showed that the addition of an
artificial thermal conductivity can dramatically improve the ability
of the SPH codes to exhibit this instability. There have since been a
number of other papers discussing this issue, but to our knowledge
none of these improvements have yet been incorporated into an SPH
model of a WD merger. Another reason for caution is that other than the
most recent results of \cite{kashyap:2015}, no white dwarf merger simulation has self-consistently
resulted in a thermonuclear detonation. Reproducibility of the detonation 
through numerical simulation is critical for building 
confidence in this progenitor model.

This paper is the first in a series designed to address these
outstanding theoretical issues for white dwarf mergers. This work will
discuss the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. \autoref{sec:Numerical Methodology}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. 
\autoref{sec:implementation} describes the method we use for setting up a
binary white dwarf simulation. \autoref{sec:Tests} discusses a few
test problems that we use to verify that our code accurately
solves the equations of fluid dynamics. \autoref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, \autoref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

To study the white dwarf merger problem, we use the grid-based
hydrodynamics code \castro\footnote{\castro\ can be obtained at \url{https://github.com/BoxLib-Codes/Castro}.} \citep{castro}.
\castro\ solves the Euler
equations, along with the inclusion of optional modules for gravity,
nuclear reactions and thermodynamics. \castro\ is based on the \boxlib
\footnote{\boxlib\ can be obtained at \url{https://github.com/BoxLib-Codes/BoxLib}.}
adaptive-mesh refinement (AMR) framework \citep{rendleman:2000}, which
represents fluid data on a mesh where regions of interest have higher
spatial resolution. \castro\ is highly parallel and is designed for
large-scale use on modern supercomputers; see 
\autoref{sec:Performance} for information on how \castro\ performs for our
problem. The next few subsections describe our approach to each of the
physics components used in this work. We direct the reader to the
original code paper for a full description of \castro's approach to
solving the equations of hydrodynamics. In this work, we report mainly
on the changes we have made to the code since its original release,
for the purpose of approaching this problem.

@@

A related issue that required a code improvement is that in cases of
large density gradients such as the edge of a star, it is possible to
generate negative densities in zones even with the more strongly
limited PPM. This can occur if a region of large density is moving
away from an ambient zone at relatively large speeds; then the net
density flux in the ambient zones can be large enough to unphysically
drag the density below zero. In practice, this will occur at the
trailing edge of a star that is moving across a grid. In such a
situation, there are two main approaches one could take: either
explicitly introduce a positivity-guaranteeing diffusive flux, or
reset the properties of the affected zone. We choose the latter
approach. Even though it is non-conservative, it preserves a
characteristic we value, which is to keep the edge of the stars
relatively sharp, as they physically should be. Since the mass of the
affected zones is typically already fairly low, this should not
seriously affect the energy conservation properties of our
simulation. Our strategy for a reset is as follows: when the density of 
a zone is below a pre-determined density floor (which is typically 
$10^{-5}\ \text{g cm}^{-3}$ for our stellar simulations), we look
at all adjacent zones and find the zone with the highest density.
If it is above the density floor, then we set the field values 
(density, momentum, energy, and temperature) of the
reset zone to be exactly equal to the field values of this 
adjacent zone. If no adjacent zone reaches the density floor, then
the zone is set to the density floor, and given a temperature equal 
to the temperature floor for our simulations (which is typically 
$10^{5}\ \text{K}$ for our stellar simulations). We then recompute 
the thermodynamics to be consistent with these values. The 
velocity of the zone is set to zero. This latter approach will only
occur in very rare situations, and is mainly there as a safeguard.

\castro's approach to adaptive mesh refinement, based on its underlying
\boxlib\ framework, is to refine zones based on certain user-specified
criteria that tag regions of interest for higher spatial
resolution. Data is represented on one of a number of AMR levels,
where each level corresponds to a set of zones at the same resolution,
which covers a subset of the domain covered by the level immediately
below it. We typically call the level 0 grid the \textit{coarse} grid,
which has the lowest spatial resolution. Each finer, higher-level grid
has a higher resolution than the grid below it by some integer factor
$N$, which is restricted to be $N = 2\ \text{or}\ 4$ in \castro. The
zones are strictly contained within the rectangular extent of the
underlying coarser zones (the code is not restricted to representing
only Cartesian geometries, but we use a Cartesian mesh with uniform
spacing in each dimension for the present study). For the time
evolution of the AMR system we use subcycling, where each AMR level is
advanced at a different timestep and a correction step is applied at
the end to synchronize the various levels. The number of
subcycled timesteps is equal to the jump in refinement between levels,
so for example on a grid with three levels and two jumps of four in
refinement, the level 2 zones will have 16 times higher spatial
resolution than the coarse grid and there will be 16 level 2 timesteps
per level 0 timestep.

@@

\castro\ has the ability to do both nuclear reactions and radiative 
transport (in the flux-limited diffusion approximation). For all simulations 
in this paper we do not enable either, and will delay discussion of 
these modules until later papers in this series.

@@

There are two ways that these source terms enter the system evolution. 
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. We choose a more accurate approach, 
which is also second order, of tracing
under the source term; the details of this are described in 
\autoref{app:hydro}. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
We believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

In a system with self-gravity, total momentum is conserved if the spatial domain
includes all of the mass of the system. This must be the 
case because each mass element exerts an equal and opposite gravitational force 
on every other mass element. However, the standard approach does not necessarily
guarantee that momentum is conserved numerically. We cannot represent a vacuum state 
in our code, so there is a small but non-zero density on the edge of the grid. 
This allows momentum to leak out of the domain even if the gravitational source term 
is written in an explicitly conservative manner. To see this, one can use the Poisson equation to write the 
density in terms of the potential and then consider its spatial discretization. For simplicity,
we will consider one spatial dimension and a uniform discretization. Analogous results 
may be readily obtained for the non-uniform case.
\begin{align}
  -\rho_{i}  \frac{d\Phi_{i}}{dx} &= -\frac{1}{4\pi G} \frac{d^2\Phi_i}{dx^2} \frac{d \Phi_i}{dx} \notag \\
  &= -\frac{1}{4\pi G} \left[\frac{\Phi_{i-1} - 2 \Phi_{i} + \Phi_{i+1}}{\Delta x^2}\right] \left[ \frac{\Phi_{i+1} - \Phi_{i-1}}{2\Delta x} \right] \notag \\
  &= -\frac{1}{8\pi G \Delta x^3} \left[ \Phi_{i+1}^2 - \Phi_{i-1}^2 - 2\Phi_i\left(\Phi_{i+1} - \Phi_{i-1}\right) \right] \label{eq:momentum_discretization}
\end{align}
It is easy to verify that adding the source terms for the current zone and the two zones 
to the left and right will result in complete cancellation of the source terms.
The catch is that if the potential if non-zero outside of the domain, then there will be
momentum lost or gained from the grid, which will be encapsulated in the ghost cells
just outside the domain. In addition, when we replace the Laplacian above by the full
three-dimensional stencil including the $y$ and $z$ derivatives, depending on the
discretization these may not be cancelled at all. This latter problem can be resolved by
writing the momentum update in an explicitly conservative way.

@@

Conservation of total energy requires that a change in gravitational energy is compensated
for by a change in gas energy, and that energy changes due to mass transfer are explicitly and 
exactly tracked. Suppose that we have some fluid mass $\Delta M_{i+1/2} = \Delta \rho_{i+1/2} \Delta V$ leave some zone
with index $i$ and enter some zone with index $i+1$. The subscript indicates that the mass change is
occurring at the interface between the two zones, at index $i+1/2$. The work done by the gravitational
force on the gas is $\Delta (\rho E) = W = \int F dx = (\Delta M_{i+1/2}\ g_{i+1/2}) (\Delta x / 2)$,
where $g_{i+1/2}$ is the gravitational acceleration at the interface. The second term in parentheses
is just the distance from the zone center to the zone edge: once the mass leaves the zone edge, it no longer
needs to be tracked. To second order, $g_{i+1/2} = -(\Phi_{i+1} - \Phi_{i}) / \Delta x$, and also to second order the potential
at the interface is given by $\Phi_{i+1/2} = (\Phi_{i+1} + \Phi_i) / 2$, so we can equivalently view the work done
as $W = -\Delta M_{i+1/2} (\Phi_{i+1/2} - \Phi_i)$. Physically, this is just the negative of the gravitational
potential energy change as the fluid is pushed from the cell center potential to the cell edge potential,
exactly as the work-energy theorem implies.

@@

In passing, the factor of $1/2$ that appears in \autoref{eq:grav_energy_conservation_update} 
has nothing to do with the factor of $1/2$ that appears in the statement of conservation of total energy, 
\autoref{eq:total_energy_gravity}. The former comes simply from the fact that the energy change is 
evaluated using the mass motion through a distance of half of the zone width. The latter is needed 
to ensure that these local changes in energy are not double-counted when doing a global integral, 
since the gravitational potential is self-generated. \autoref{eq:grav_energy_conservation_update} 
applies to any conservative potential $\Phi$, and we will use this to our advantage for the 
rotation forces in \autoref{sec:rotation}.

@@

Another approach to conserving total energy recently taken in the literature is to evolve an 
equation for the total energy $(\rho E_{\text{tot}})$; see \cite{marcello:2012} and \cite{jiang:2013} (see also 
\cite{arepo}, Section 5.3). That is, one can replace the gas energy equation with a total energy equation, 
and then the energy flux includes a term corresponding to the flux of gravitational potential energy. We 
avoid this approach for our problem because there will be regions on the computational domain where the total 
energy is dominated by potential energy (especially the low-density regions near the edge of the white dwarfs),
and the gas energy can only be retrieved by first subtracting $-\rho \Phi/2$ from the total energy. Like 
\cite{arepo}, we find that this can result in some serious errors due to numerical discretization, yielding 
unphysical energies or temperatures. We observe also that the implementation of \cite{jiang:2013} 
has terms in the gravitational flux that are not proportional to $\rho$, and so can lead to the 
same troubles that plague the tensor-based formalism for the momentum equation, where small errors 
in the discretization of the gravitational potential can lead to very large changes in the energy of the gas.

@@

This approach becomes computationally feasible when we cut off the
outer summation in \autoref{eq:multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions will be directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.

>>>>

\documentclass[iop,numberedappendix]{../emulateapj}

@@

\author{Max P. Katz, Michael Zingale, Alan C. Calder, F. Douglas Swesty, Ann S. Almgren, Weiqun Zhang}
%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
The Type Ia supernova progenitor problem is one of the most perplexing and 
exciting problems in astrophysics, requiring detailed numerical modeling to 
complement the observations of these explosions. One possible progenitor that has 
merited recent theoretical attention is the white dwarf merger scenario.
This can naturally explain many of the observed characteristics of 
Type Ia supernovae, as well as their distribution in time and space.
To date there have been relatively few fully self-consistent simulations 
of merging binary white dwarf systems using mesh-based hydrodynamics. 
This is the first paper in a series designed to 
describe simulations of these systems using a mesh-based hydrodynamics code with 
adaptive mesh refinement. In this paper we describe our numerical 
methodology and discuss our implementation in the code \castro. \castro\ 
solves the Euler equations, and for merging white dwarf simulations we use its capability to solve 
the Poisson equation for self-gravity and couple the gravitational force to the hydrodynamics.
Standard techniques for coupling gravitation and rotation forces to the hydrodynamics do 
not do an adequate job conserving the total energy of the system for our problem, 
but recent advances in the literature allow progress on this 
issue and we discuss our implementation here. We present a
set of test problems demonstrating the extent to which our software sufficiently
models a system where large amounts of mass are advected on the computational 
domain over very long timescales. Future papers in this series will describe
our treatment of the initial conditions of these systems and will 
examine the early phases of the merger to determine its viability
for triggering a thermonuclear detonation.

@@

Around the same time, however, \cite{guillochon:2010} and
\cite{dan:2011} pointed out that the previously mentioned simulations 
generally shared a significant drawback, which was that their initial conditions
were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006},
and \cite{motl:2007} (the first three-dimensional mesh-based
simulations of mass transfer in binary white dwarf systems) pioneered
the study of the long-term dynamical evolution of binary
white dwarf systems after constructing equilibrium initial
conditions. Earlier work placed the stars too close together 
and ignored the effects of tidal forces that change the shape of the 
secondary, leading to the merger
happening artificially too quickly \citep{fryer:2008}. When the initial conditions are
constructed in equilibrium, the system can be stable for tens of
orbital periods, substantially changing the character of the mass
transfer phase. One limitation of this series of studies is
that the authors used a polytropic equation of state and thus could
not consider nuclear reactions. \cite{guillochon:2010} and
\cite{dan:2011} improved on this using a realistic equation of state,
a nuclear reaction network, and a similar approach to the equilibrium
initial conditions, and found substantial agreement with the idea that
mass transfer occurs in a stable manner over tens of orbital
periods. They also found that, assuming the material accreted onto the
surface of the primary was primarily helium, explosive surface
detonations would occur as a result of accretion stream instabilities
during the mass transfer phase prior to the full merger. This could
trigger a double-detonation explosion and thus perhaps a SN Ia.

The latest developments have resulted in some possible areas of convergence.
\cite{pakmor:2012} performed a merger scenario
with a $1.1\ \msolar$ and $0.9\ \msolar$ setup, with better treatment
of the initial conditions, and indeed found that the merger process
happened over more than ten orbits. Nevertheless, they still determined
that a carbon-oxygen detonation would occur, in line with their
earlier results. \cite{moll:2014} and \cite{kashyap:2015} were also 
able to find a detonation in similarly massive systems. Notably,
the detonation occurred self-consistently and did not need to be  
intentionally triggered using an external source term.
\cite{dan:2012} and \cite{dan:2014} performed a large sweep 
of the parameter space for merger pairs and
found that pure carbon-oxygen systems would generally not lead to
detonations and violent mergers except for the most massive
systems. They did find that for systems containing helium, many
would detonate and potentially lead to SNe Ia, either through the
aforementioned instabilities in the accretion stream, or during the
contact phase, similar to the violent carbon-oxygen
mergers. \cite{pakmor:2013} added a thin helium shell on their primary
white dwarf, and found that this robustly led to a detonation of the
white dwarf. For now there is preliminary support for the hypothesis
that systems with helium shells (or helium WDs), and very massive carbon-oxygen binaries,
could robustly lead to events resembling SNe Ia.

Given the considerable research into the double degenerate problem 
described above, why is another approach using a different simulation
code warranted? First and foremost, reproducibility of the results
across simulation codes and algorithms is important for gauging
confidence in this result. Most of the existing results that study 
the viability of double degenerate systems as progenitors for
Type Ia supernovae (that is, including a realistic 
equation of state and nuclear reactions) have
used the SPH method. SPH codes have a number of features which do aid
them in the study of these systems, such as conservation of
angular momentum to machine precision when there are no source terms
such as gravity (when gravity is used, conservation is limited by the
tolerance of error in the gravity solver). A drawback relates to the fact that whether a prompt detonation
in a merger happens depends in detail on the nature of the
gas at the interface between the two stars, which is at much lower
density than the rest of the stellar material. The SPH codes use
uniform mass particles, so their effective resolution is
\textit{lowest} at the stellar surface. In contrast, a code
with adaptive mesh refinement can zoom in on the regions where
hotspots will develop, while also maintaining high enough resolution
in the high-density regions to adequately capture the large-scale mass
transfer dynamics. There are also outstanding questions of
convergence in SPH (e.g.\ \citealt{zhu_SPH:2014}) and whether the method
correctly captures fluid instabilities. This is an important question
for white dwarf mergers because of the likely importance small-scale
instabilities will have on the evolution of the low-density gas at the
primary's surface. The pioneering work of \cite{agertz:2007} compared
grid and SPH codes and found some important differences. Most relevant
for this discussion is that the SPH codes could not adequately handle
mixing from the Kelvin-Helmholtz instability in the test they
propose. As pointed out by \cite{price:2008}, this is not a result of
SPH being inherently unable to model this instability, but instead it
is attributed to the fact that the standard SPH evolution equations do
not have a mechanism for capturing discontinuities in internal
energy. \citeauthor{price:2008} showed that the addition of an
artificial thermal conductivity can dramatically improve the ability
of the SPH codes to exhibit this instability. There have since been a
number of other papers discussing this issue, but to our knowledge
none of these improvements have yet been incorporated into an SPH
model of a WD merger. Another reason for caution is that other than the
most recent results of \cite{kashyap:2015}, no white dwarf merger simulation has self-consistently
resulted in a thermonuclear detonation. Reproducibility of the detonation 
through numerical simulation is critical for building 
confidence in this progenitor model.

This paper is the first in a series designed to address these
outstanding theoretical issues for white dwarf mergers. This work 
discusses the verification of our hydrodynamics code for simulating
these events. Later efforts will look at the initial conditions of the
system, the robustness with which a hotspot is found from which a
detonation could occur, and the importance of the initial white dwarf
models, which should be more sophisticated than simple carbon-oxygen
mixtures and in principle should use results from modern stellar
evolution calculations. \autoref{sec:Numerical Methodology}
describes our code and why it can provide useful results compared to
other methodologies used for this problem. 
\autoref{sec:implementation} describes the method we use for setting up a
binary white dwarf simulation. \autoref{sec:Tests} discusses a few
test problems that we use to verify that our code accurately
solves the equations of fluid dynamics. \autoref{sec:Performance}
demonstrates that the software scales well for supercomputer
applications. Finally, \autoref{sec:Conclusions and Discussion}
recaps what we have shown and highlights some of the future work we
plan to do.

@@

To study the white dwarf merger problem, we use the mesh-based
hydrodynamics code \castro\footnote{\castro\ can be obtained at \url{https://github.com/BoxLib-Codes/Castro}.} \citep{castro}.
\castro\ solves the Euler
equations, along with the inclusion of optional modules for gravity,
nuclear reactions and thermodynamics. \castro\ is based on the \boxlib
\footnote{\boxlib\ can be obtained at \url{https://github.com/BoxLib-Codes/BoxLib}.}
adaptive-mesh refinement (AMR) framework \citep{rendleman:2000}, which
represents fluid data on a hierarchical mesh where regions of interest have higher
spatial resolution. \castro\ is highly parallel and is designed for
large-scale use on modern supercomputers; see 
\autoref{sec:Performance} for information on how \castro\ performs for our
problem. The next few subsections describe our approach to each of the
physics components used in this work. We direct the reader to the
original code paper for a full description of \castro's approach to
solving the equations of hydrodynamics. In this work, we report mainly
on the changes we have made to the code since its original release,
for the purpose of approaching this problem.

@@

A related issue that required a code improvement is that in cases of
large density gradients such as the edge of a star, it is possible to
generate negative densities in zones even with the more strongly
limited PPM. This can occur if a region of large density is moving
away from an ambient zone at relatively large speeds; then the net
density flux in the ambient zones can be large enough to unphysically
drag the density below zero. In practice, this occurs at the
trailing edge of a star that is moving across a grid. In such a
situation, there are two main approaches one could take: either
explicitly introduce a positivity-guaranteeing diffusive flux, or
reset the properties of the affected zone. We choose the latter
approach. Even though it is non-conservative, it preserves a
characteristic we value, which is to keep the edge of the stars
relatively sharp, as they physically should be. Since the mass of the
affected zones is typically already fairly low, this should not
seriously affect the energy conservation properties of our
simulation. Our strategy for a reset is as follows: when the density of 
a zone is below a pre-determined density floor (which is typically 
$10^{-5}\ \text{g cm}^{-3}$ for our stellar simulations), we look
at all adjacent zones and find the zone with the highest density.
If it is above the density floor, then we set the field values 
(density, momentum, energy, and temperature) of the
reset zone to be exactly equal to the field values of this 
adjacent zone. If no adjacent zone reaches the density floor, then
the zone is set to the density floor, and given a temperature equal 
to the temperature floor for our simulations (which is typically 
$10^{5}\ \text{K}$ for our stellar simulations). We then recompute 
the thermodynamics to be consistent with these values. The 
velocity of the zone is set to zero. This latter approach only
occurs in very rare situations, and is mainly there as a safeguard.

\castro's approach to adaptive mesh refinement, based on its underlying
\boxlib\ framework, is to refine zones based on certain user-specified
criteria that tag regions of interest for higher spatial
resolution. Data is represented on one of a number of AMR levels,
where each level corresponds to a set of zones at the same resolution,
which covers a subset of the domain covered by the level immediately
below it. We typically call the level 0 grid the \textit{coarse} grid,
which has the lowest spatial resolution. Each finer, higher-level grid
has a higher resolution than the grid below it by some integer factor
$N$, which is restricted to be $N = 2\ \text{or}\ 4$ in \castro. The
zones are strictly contained within the rectangular extent of the
underlying coarser zones (the code is not restricted to representing
only Cartesian geometries, but we use a Cartesian mesh with uniform
spacing in each dimension for the present study). For the time
evolution of the AMR system we use subcycling, where each AMR level is
advanced at a different timestep and a correction step is applied at
the end to synchronize the various levels. The number of
subcycled timesteps is equal to the jump in refinement between levels,
so for example on a grid with three levels and two jumps of four in
refinement, the level 2 zones have 16 times higher spatial
resolution than the coarse grid and there are 16 level 2 timesteps
per level 0 timestep.

@@

\castro\ has the ability to do both nuclear reactions and radiative 
transport (in the flux-limited diffusion approximation). For all simulations 
in this paper we do not enable either, and we delay discussion of 
these modules until later papers in this series.

@@

There are two ways that these source terms enter the system evolution. 
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. We choose a more accurate approach, 
which is also second order, of tracing
under the source term; the details of this are described in 
\autoref{app:hydro}. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we describe how we do this for the momentum,
and then we describe our approach for the energy. This discussion is somewhat detailed.
We believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

In a system with self-gravity, total momentum is conserved if the spatial domain
includes all of the mass of the system. This must be the 
case because each mass element exerts an equal and opposite gravitational force 
on every other mass element. However, the standard approach does not necessarily
guarantee that momentum is conserved numerically. We cannot represent a vacuum state 
in our code, so there is a small but non-zero density on the edge of the grid. 
This allows momentum to leak out of the domain even if the gravitational source term 
is written in an explicitly conservative manner. To see this, one can use the Poisson equation to write the 
density in terms of the potential and then consider its spatial discretization. For simplicity,
we consider one spatial dimension and a uniform discretization. Analogous results 
may be readily obtained for the non-uniform case.
\begin{align}
  -\rho_{i}  \frac{d\Phi_{i}}{dx} &= -\frac{1}{4\pi G} \frac{d^2\Phi_i}{dx^2} \frac{d \Phi_i}{dx} \notag \\
  &= -\frac{1}{4\pi G} \left[\frac{\Phi_{i-1} - 2 \Phi_{i} + \Phi_{i+1}}{\Delta x^2}\right] \left[ \frac{\Phi_{i+1} - \Phi_{i-1}}{2\Delta x} \right] \notag \\
  &= -\frac{1}{8\pi G \Delta x^3} \left[ \Phi_{i+1}^2 - \Phi_{i-1}^2 - 2\Phi_i\left(\Phi_{i+1} - \Phi_{i-1}\right) \right] \label{eq:momentum_discretization}
\end{align}
It is easy to verify that adding the source terms for the current zone and the two zones 
to the left and right results in complete cancellation of the source terms.
The catch is that if the potential if non-zero outside of the domain, then there will be
momentum lost or gained from the grid, which will be encapsulated in the ghost cells
just outside the domain. In addition, when we replace the Laplacian above by the full
three-dimensional stencil including the $y$ and $z$ derivatives, depending on the
discretization these may not be cancelled at all. This latter problem can be resolved by
writing the momentum update in an explicitly conservative way.

@@

Conservation of total energy requires that a change in gravitational energy is compensated
for by a change in gas energy, and that energy changes due to mass transfer are explicitly and 
exactly tracked. Suppose that we have some fluid mass $\Delta M_{i+1/2} = \Delta \rho_{i+1/2} \Delta V$ leave the zone
with index $i$ and enter some zone with index $i+1$. The subscript indicates that the mass change is
occurring at the interface between the two zones, at index $i+1/2$. The work done by the gravitational
force on the gas is $\Delta (\rho E) = W = \int F dx = (\Delta M_{i+1/2}\ g_{i+1/2}) (\Delta x / 2)$,
where $g_{i+1/2}$ is the gravitational acceleration at the interface. The second term in parentheses
is just the distance from the zone center to the zone edge: once the mass leaves the zone edge, it no longer
needs to be tracked. To second order, $g_{i+1/2} = -(\Phi_{i+1} - \Phi_{i}) / \Delta x$, and also to second order the potential
at the interface is given by $\Phi_{i+1/2} = (\Phi_{i+1} + \Phi_i) / 2$, so we can equivalently view the work done
as $W = -\Delta M_{i+1/2} (\Phi_{i+1/2} - \Phi_i)$. Physically, this is just the negative of the gravitational
potential energy change as the fluid is pushed from the cell center potential to the cell edge potential,
exactly as the work-energy theorem implies.

@@

In passing, the factor of $1/2$ that appears in \autoref{eq:grav_energy_conservation_update} 
has nothing to do with the factor of $1/2$ that appears in the statement of conservation of total energy, 
\autoref{eq:total_energy_gravity}. The former comes simply from the fact that the energy change is 
evaluated using the mass motion through a distance of half of the zone width. The latter is needed 
to ensure that these local changes in energy are not double-counted when doing a global integral, 
since the gravitational potential is self-generated. \autoref{eq:grav_energy_conservation_update} 
applies to any conservative potential $\Phi$, and we use this to our advantage for the 
rotation forces in \autoref{sec:rotation}.

@@

Another approach to conserving total energy recently taken in the literature is to evolve an 
equation for the total energy $(\rho E_{\text{tot}})$; see \cite{marcello:2012} and \cite{jiang:2013} (see also 
\cite{arepo}, Section 5.3). That is, one can replace the gas energy equation with a total energy equation, 
and then the energy flux includes a term corresponding to the flux of gravitational potential energy. We 
avoid this approach for our problem because there are regions on the computational domain where the total 
energy is dominated by potential energy (especially the low-density regions near the edge of the white dwarfs),
and the gas energy can only be retrieved by first subtracting $-\rho \Phi/2$ from the total energy. Like 
\cite{arepo}, we find that this can result in some serious errors due to numerical discretization, yielding 
unphysical energies or temperatures. We observe also that the implementation of \cite{jiang:2013} 
has terms in the gravitational flux that are not proportional to $\rho$, and so can lead to the 
same troubles that plague the tensor-based formalism for the momentum equation, where small errors 
in the discretization of the gravitational potential can lead to very large changes in the energy of the gas.

@@

This approach becomes computationally feasible when we cut off the
outer summation in \autoref{eq:multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions is directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.
